---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cycif.importer

<!-- badges: start -->
<!-- badges: end -->

The `cycif.importer` package provides functions to import CyCIF
(Cyclic Immunofluorescence) multiplexed tissue imaging data ran through the
[MCMICRO](https://mcmicro.org/) pipeline into R and perform basic ROI and marker
gating.

## Features

- **Data Import**: Load UNMICST cell segmentation data and ROI definitions
- **Gating**: Apply single and double marker gating thresholds
- **ROI Assignment**: Assign cells to regions of interest polygons
- **Marker Combinations**: Phenotype cells using custom marker combinations
- **Sampling**: Subsample cells
- **Summarization**: Generate stratified summaries with custom groupings

## Installation

You can install the development version of cycif.importer from [GitHub](https://github.com/clemenshug/cycif.importer) with:

``` r
# install.packages("remotes")
remotes::install_github("clemenshug/cycif.importer")
```

## Pipeline Overview

The `cycif.importer` package implements a modular three-stage analysis pipeline:

### Stage 1: Import & Gate
- Load cell quantification data from UNMICST output
- Load ROI definitions
- Apply single and double marker gates
- Assign cells to ROIs with optional boundary expansion

### Stage 2: Sample
- Apply flexible sampling strategies
- Support for all-cells or ROI-only sampling
- Configurable sample sizes per slide

### Stage 3: Summarize
- Generate summary statistics with custom filters
- Support for multiple grouping strategies
- Export results as CSV files

## Example with Built-in Data

The package includes example data from two tissue samples with pre-defined gates and ROIs:

```{r detailed-example, echo=TRUE, message=FALSE, warning=FALSE}
library(cycif.importer)
library(dplyr)

# Set up paths to example data
data_dir <- system.file("extdata/example_quants", package = "cycif.importer")
roi_dir <- system.file("extdata/example_rois", package = "cycif.importer")
gate_file <- system.file("extdata/example_gates.csv", package = "cycif.importer")
output_dir <- tempdir()

# Run the complete pipeline
results <- cycif_pipeline(
  counts = data_dir,
  rois = roi_dir,
  gate_table = gate_file,
  output_dir = output_dir,
  summary_filters = list(
    `PanCK+` = ~PanCKp,
    `PanCK-` = ~!PanCKp,
    `CD8+` = ~CD8ap,
    `CD4+` = ~CD4p
  ),
  summary_groups = list(
    by_roi = c("slideName", "ROI", "ROIname"),
    by_slide = c("slideName"),
    by_roi_type = c("slideName", "ROIname")
  ),
  sample_size = 25000,
  sampling_mode = "all_cells",
  expand_distance = 50  # Expand ROI boundaries by 50 microns
)
```

The `output_dir` directory will contain the following files:

- `sampled_cells.csv.gz`: Cells sampled from the dataset with gates and ROI assignments
- Summaries for each of the defined filter and grouping combinations e.g.
  - `by_roi__PanCK+.csv.gz`: Summary statistics grouped by ROI including only PanCK+ cells
  - `by_slide__CD8+.csv.gz`: Summary statistics grouped by slide including only CD8+ cells

```{r show-files, echo=FALSE}
list.files(output_dir)
```

## Modular Usage

You can also use individual functions for more control:

```{r modular-example, eval=FALSE}
# Load data step by step
cell_data <- cycif_load_cell_data(data_dir)
roi_data <- cycif_load_roi_data(roi_dir, names(cell_data))

# Apply gates
gates <- readr::read_csv(gate_file)
gated_data <- cycif_apply_gates(cell_data, gates)

# Assign ROIs
data_with_rois <- cycif_assign_rois(
  gated_data,
  roi_data,
  scale_factor = 0.65,
  expand_distance = 50
)

# Sample cells
sampled_data <- cycif_sample(
  data_with_rois,
  sample_size = 10000,
  sampling_mode = "roi_only"
)

# Generate summaries
summaries <- cycif_summarize(
  sampled_data,
  summary_filters = list(`CD8+` = ~CD8ap),
  summary_groups = list(by_slide = "slideName")
)
```

## Data Formats

The package expects specific input file formats. Here are examples using the built-in test data:

### Cell Quantification Data (UNMICST Output)

Cell-level quantification files from UNMICST segmentation (`.csv` or `.csv.gz` format)

#### Directory Structure

The cell quantification data directory should contain one `.csv` or `.csv.gz` file per slide:

```
data_dir/
├── LSP11060--unmicst_cell.csv.gz
├── LSP11064--unmicst_cell.csv.gz
└── LSP11065--unmicst_cell.csv
```

- Files should follow the pattern: `{slideName}--unmicst_cell.csv[.gz]`
- The `slideName` portion (before `--unmicst_cell`) must match the `slideName` in your gate thresholds file
- Compressed (`.gz`) files are automatically detected and handled

#### Example

```{r cell-data-example, echo=FALSE, message=FALSE, warning=FALSE}
# Load actual example cell data
cell_file <- system.file("extdata/example_quants/LSP11060--unmicst_cell.csv.gz",
                        package = "cycif.importer")
if (file.exists(cell_file)) {
  cell_data <- readr::read_csv(cell_file, show_col_types = FALSE)
  # Show first few rows and selected columns
  knitr::kable(
    cell_data[1:3, c("CellID", "X_centroid", "Y_centroid", "Area", "PanCK", "CD8a", "CD4", "FoxP3")],
    caption = "Example Cell Quantification Data",
    digits = 2
  )
} else {
  cat("Cell data file not found for display")
}
```

**Required columns**:

- `CellID`: Unique cell identifier
- `X_centroid`, `Y_centroid`: Cell center coordinates (pixels)
- **Marker columns**: Quantified intensity values for each marker

**Optional columns**:

- `Area`: Cell area
- `MajorAxisLength`, `MinorAxisLength`: Cell shape metrics
- Additional morphological features

### Gate Thresholds File

CSV file containing marker intensity thresholds for each slide:

```{r gates-example, echo=FALSE, message=FALSE, warning=FALSE}
# Load actual example gate data
gate_file <- system.file("extdata/example_gates.csv", package = "cycif.importer")
if (file.exists(gate_file)) {
  gates <- readr::read_csv(gate_file, show_col_types = FALSE)
  # Display available columns
  available_cols <- intersect(c("slideName", "PanCK", "CD8a", "CD4", "FoxP3", "Ki67"), names(gates))
  if (length(available_cols) > 1) {
    knitr::kable(
      gates[, available_cols],
      caption = "Example Gate Thresholds (log2 transformed intensities)",
      digits = 2
    )
  } else {
    knitr::kable(
      gates[, seq_len(min(6, ncol(gates)))],
      caption = "Example Gate Thresholds (log2 transformed intensities)",
      digits = 2
    )
  }
} else {
  cat("Gate thresholds file not found for display")
}
```

**Required columns**:

- `slideName`: Must match slide names in cell data files
- **Marker columns**: Log2-transformed intensity thresholds for each marker

### Region of Interest (ROI) Files

CSV files defining polygon regions for each slide.

#### Directory Structure
The ROI directory should contain one `.csv` file per slide with ROI definitions:

```
roi_dir/
├── LSP11060-rois.csv
├── LSP11064-rois.csv
└── LSP11065-rois.csv
```

- Files should follow the pattern: `{slideName}-rois.csv`
- The `slideName` portion (before `-rois`) must match the slide names in your cell quantification files
- Each file contains all ROI polygons for that specific slide

#### Example

```{r roi-example, echo=FALSE, message=FALSE, warning=FALSE}
# Load actual example ROI data
roi_file <- system.file("extdata/example_rois/LSP11060-rois.csv", package = "cycif.importer")
if (file.exists(roi_file)) {
  rois <- readr::read_csv(roi_file, show_col_types = FALSE)
  # Show first few rows and selected columns if they exist
  available_cols <- intersect(c("Id", "Name", "Text", "type"), names(rois))
  if (length(available_cols) > 0) {
    knitr::kable(
      rois[seq_len(min(3, nrow(rois))), available_cols],
      caption = "Example ROI Definitions"
    )
  } else {
    # Show first few columns if specific ones aren't available
    knitr::kable(
      rois[seq_len(min(3, nrow(rois))), seq_len(min(4, ncol(rois)))],
      caption = "Example ROI Definitions"
    )
  }
} else {
  cat("ROI file not found for display")
}
```

**Required columns**:

- `Id`: Unique ROI identifier
- `Name`: ROI name/identifier
- `all_points`: Polygon coordinates as "x1,y1 x2,y2 ..." string

**Optional columns**:

- `Text`: ROI description
- Geometric properties (X, Y, Width, Height, etc.)

### Slide Metadata (Optional)

CSV file with additional slide-level information:

```{r slide-meta-example, echo=FALSE, message=FALSE, warning=FALSE}
# Create example slide metadata structure based on available slides
slide_names <- c("LSP11060", "LSP11064")  # From the example data
slide_meta <- data.frame(
  slideName = slide_names,
  patient_id = c("P001", "P002"),
  tissue_type = c("Ovary", "Fallopian tube"),
  treatment = c("Control", "Treated"),
  stringsAsFactors = FALSE
)

knitr::kable(slide_meta, caption = "Example Slide Metadata Structure")
```

**Required columns**:

- `slideName`: Must match slide names in other files

**Optional columns**:

- Any additional slide-level annotations

### ROI Metadata (Optional)

CSV file with additional ROI-level information:

```{r roi-meta-example, echo=FALSE, message=FALSE, warning=FALSE}
# Create example ROI metadata structure
roi_meta <- data.frame(
  ROIname = c("Incidental STIC", "Incidental Fimbriae", "Incidental STIL"),
  region_type = c("Lesion", "Normal", "Lesion"),
  pathologist_score = c(3, 1, 2),
  stringsAsFactors = FALSE
)

knitr::kable(roi_meta, caption = "Example ROI Metadata Structure")
```

**Required columns**:

- `ROIname`: Must match ROI names in ROI definition files

**Optional columns**:

- Any additional ROI-level annotations

### Output Data

- **Sampled Cells**: Cell-level data with gates, ROI assignments, and metadata
- **Summaries**: Aggregated statistics by user-defined groups and filters

## Customization

The pipeline supports extensive customization:

```{r customization, eval=FALSE}
# Custom marker combinations
marker_combos <- list(
  regulatory_T = ~CD4p & FoxP3p,
  cytotoxic_T = ~CD8ap & GZMBp,
  exhausted_T = ~CD8ap & PD1p
)

# Custom double gates
double_gates <- data.frame(
  marker1 = c("CD4", "CD8a"),
  marker2 = c("FoxP3", "GZMB")
)

# Run with custom parameters
results <- cycif_pipeline(
  data_dir = data_dir,
  roi_dir = roi_dir,
  marker_combinations = marker_combos,
  double_gates = double_gates,
  scale_factor = 0.5,  # Different pixel-to-micron conversion
  expand_distance = 100,  # Larger ROI expansion
  output_dir = output_dir
)
```
